---
title: "Module 3-1-Principle - Data Visualization with ggplot2 in R"
author: "Angelina Jao"
date: '`r format(Sys.time())`'
format: 
  html: 
    toc: true
    toc-depth: 4
    toc-location: right-body
    number-sections: false
    code-line-numbers: true
    code-fold: false
    code-link: true
    embed-resources: true
    df-print: paged
editor: visual
execute: 
  freeze: auto
  warning: false  
  error: true
---

# Overview

## Expected Learning Outcomes

> After taking this workshop, participants should be able to do the following:

1.  Explain the concept of grammar of graphics when visualizing data with the ggplot2 package.
2.  Be familiar with various types of charts.
3.  Visualize data in counts and proportions.
4.  Select appropriate charts based on strategic considerations (e.g., the characteristics of the data and the audience).
5.  Create a chart that involves one or two variables with either categorical or continuous data.
6.  Create a chart by adding a categorical moderator (3rd variable) to the chart involving two or three variables.
7.  Create correlation charts.
8.  Read charts and generate insights.
9.  Describe three popular packages that allow one to visualize data.
10. Explain the concept of the grammar of graphics when visualizing data with the ggplot2 package.

## Loading Packages

```{r}
#install.packages("ggplot2")
#install.packages('ggrepel')
#install.packages('ggthemes')
#install.packages('scales')
#install.packages('plotly')
#install.packages('lattice')
#install.packages('GGally')
#install.packages("dplyr")
#install.packages("tidyverse")
#install.packages("ggtext")
#install.packages("glue")
library(ggplot2) #visualization
library(ggrepel) #labels for data
library(ggthemes) #collections of themes
library(scales) # scale
library(plotly) # interactive chart
library(GGally) # correlation
library(dplyr) # data transformation
library(tidyverse) # mega package containing 8 packages
library(ggtext) # for text visualization
library(glue) # combining multiple component
library(gapminder)
```

# 1. Understand mtcars data

## 1.1 Using Help

```{r mtcars data}
#?mtcars
```

> A data frame with 32 observations on 11 (numeric) variables.

-   \[, 1\] mpg: Miles/(US) gallon
-   \[, 2\] cyl: Number of cylinders
-   \[, 3\] disp: Displacement (cu.in.)
-   \[, 4\] hp: Gross horsepower
-   \[, 5\] drat: Rear axle ratio
-   \[, 6\] wt: Weight (1000 lbs)
-   \[, 7\] qsec: 1/4 mile time
-   \[, 8\] vs: Engine (0 = V-shaped, 1 = straight)
-   \[, 9\] am: Transmission (0 = automatic, 1 = manual)
-   \[,10\] gear: Number of forward gears
-   \[,11\] carb: Number of carburetors

## 1.2 Reading data and converting to a tibble (cars)

-   short cut for adding r code chunk: ctrl + alt + I
-   short cut for executing a code ("running a code") is ctrl + enter simutaneously

```{r preparing data}
mtcars
class(mtcars)

cars <- mtcars |> 
  rownames_to_column() |>
  as_tibble() |>
  rename(model = rowname) |>
  print(n = 20, width = Inf)

cars
```

## 1.3 Simple Descriptive Statistics

```{r}
cars |>
  summary()

glimpse(cars) # base R

# tdyverse
cars |>
  glimpse()
skim(cars)
skimr::skim(cars)
```

# 2. Basic Plotting Methods in Base R

```{r}
hist(cars$disp, breaks = 10)
```

# 3 Lattice package

```{r}
# install.packages("lattice")
library(lattice)

xyplot(mpg ~ wt, data = cars)

histogram(~ disp, data = cars, breaks = 10)
```

# 4. ggplot2

-   We will use ggplot2 - the best tool in the market for data visualization - from now on.

## 4.1 Elaborate Examples

### 4.1.1. X & Y are both continous with moderator and labeller()

@complex shows an elaborate coding example with ggplot2.

```{r}
#| label: complex
cars  |>
  count(cyl)

cars |>
  mutate(cyl = factor(cyl))

easy_labels <- c("4" = "4 Cylinder Cars",
                 "6" = "6 Cylinder Cars",
                 "8" = "8 Cylinder Cars"
                 )

cars %>%
  mutate(cyl = factor(cyl)) %>%
  ggplot(aes(x = mpg, y = disp, color = cyl)) +
  geom_point(size=3,
             color='black'
             ) + #geom
  # geom_jitter() +
  geom_smooth(method = lm) +
  # facet_grid(cols = vars(cyl),
  facet_wrap(~ cyl, 
             #scales = "free_y",
             ncol = 1,
             strip.position = "top",
             labeller = labeller(cyl = easy_labels)
             ) + #faceting
  scale_y_continuous(limits = c(0, NA), expand= c(0,0)) +
  coord_flip() + #coordinate
  theme_economist() + #labels
  labs(title = 'MPG vs Displacement',
       x = 'Miles Per Gallon',
       y = 'Displacement') +
  theme(
    strip.placement = "outside",
    strip.background = element_blank(),
    panel.background = element_blank(),
    panel.grid = element_blank(),
    axis.line = element_line()
    ) +
  guides(color = 'none')
```

### 4.1.2 X & Y are both continous with moderator & as_labeller()

```{r application}
easy_labels_n <- as_labeller(c(`4` = "4 Cylinder Cars",
                               `6` = "6 Cylinder Cars",
                               `8` = "8 Cylinder Cars"
                             )
                          )

ggplot(data = cars, aes(x = disp, y = mpg, color = factor(cyl))) + #data
  geom_point(size=3) + #geometry
  facet_grid(~ factor(cyl), 
             labeller = easy_labels_n
             ) + #faceting
  theme_bw() + #theme type
  labs(title = 'MPG vs Displacement', #labels
       x = 'Displacement',
       y = 'Miles Per Gallon',
       color = "# of Cylinder") + 
  guides(color = 'none')

```

## 4.2. One continous variable: geom_histogram()

```{r}
ggplot(data = cars, aes(x = mpg)) +
  geom_histogram(binwdith = 5, color = "red", fill = "blue")

ggplot(data = cars, aes(x = mpg, fill = mpg)) +
  geom_histogram(binwidth = 3, color = "red", fill = "blue")
```

## 4.3 One categorical variable: geom_bar()

### 4.3.1. Bar chart

```{r}
barplot(table(cars$cyl)) # from graphics package

# the same as above in ggplot2
ggplot(data = cars, aes(x = factor(cyl))) + 
  geom_bar() + 
  guides(fill = "none") # remove fill legend.
```

### 4.3.2. How to reorder bars in Barplot

#### 4.3.2.1. mpg data

```{r mpg data}
#help(mpg)
```

A data frame with 233 rows and 11 variables:

-   manufacturer: manufacturer name
-   model: model name
-   displ: engine displacement, in litres
-   year: year of manufacture
-   cyl: number of cylinders
-   trans: type of transmission
-   drv: the type of drive train, where f = front-wheel drive, r = rear wheel drive, 4 = 4wd
-   cty: city miles per gallon
-   hwy: highway miles per gallon
-   fl: fuel type
-   class: "type" of car

```{r}
mpg <- mpg
head(mpg)
```

#### 4.3.2.2. Reorder manufacturers by Count of cars

```{r reordering}
ggplot(data = mpg, aes(x = manufacturer, fill = manufacturer)) + #not sorted
  geom_bar(stat = "count", show.legend = FALSE) +
  coord_flip()

# use reorder bars in inside ggplot function
mpg %>%
  count(manufacturer, sort = TRUE) %>%
  ggplot(aes(x = reorder(manufacturer, n), y = n, fill = manufacturer)) +
#reorder() or fct_reorder()/fct_rev()
  geom_bar(stat = "identity", show.legend = FALSE) +
  coord_flip()

# Use reorder bars inside mutate function during wrangling
mpg %>%
  count(manufacturer, sort = TRUE) %>%
mutate(manufacturer = fct_reorder(manufacturer, n)
       ) %>%
  ggplot(aes(x = manufacturer, y = n, fill = manufacturer)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  coord_flip()

```

## 4.4. Two Variables

### 4.4.1. Cutting a Categorical Var by a Categorical Var: Stacked vs. Dodged vs. Filled Barplots

#### 4.4.1.1. Stack vs. dodged barplots

```{r}
ggplot(data = cars, aes(x = factor(cyl), fill = factor(am))) +
  geom_bar(stat = "count", position = "stack") +
  labs(x = "Cylinder Size",
       y = "# of Cars",
       fill = "Automatic vs. Manual"
       )

ggplot(data = cars, aes(x = factor(cyl), fill = factor(am))) +
  geom_bar(position = "dodge") +
  labs(x = "Cylinder Size",
       y = "# of Cars",
       fill = "Automatic vs. Manual"
       )
```

#### 4.4.1.2. Stakced vs. Filled barplots

```{r}
# stack.
ggplot(data = mpg, aes(x = manufacturer, fill = class)) +
  geom_bar(stat = "count", position = "stack") +
  theme_economist() +
  theme(axis.text.x = element_text(angle = 90, hjust = 0)) +
  labs(title = "# of Cars by Class for Each Manufacturer",
       x = NULL,
       y = "# of cars"
             ) +
  coord_flip()

# fill.
ggplot(data = mpg, aes(x = manufacturer, fill = class)) +
  geom_bar(stat = "count", position = "fill") + #fill converts count to fraction.
  scale_y_continuous(labels = percent) +
  theme_economist() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0)) +
  labs(title = "Percent of cars by class for each Manufacturer",
       x = "",
       y = NULL
       ) +
  coord_flip()
```

#### 4.4.1.3. faceted barplots

```{r}
# manufacturer by class
ggplot(data = mpg, aes(x = manufacturer, fill = class)) +
  geom_bar(stat = "count") +
  facet_wrap(~ class) +
  theme_bw() +
  labs(title = "# of Cars for Manufacturer by class",
       x = NULL,
       y = "# of cars",
              ) +
  coord_flip() +
  guides(fill = "none")

# class by manufacturer
ggplot(data = mpg, aes(x = class, fill = manufacturer)) +
  geom_bar(stat = "count") +
  facet_wrap(~ manufacturer) +
  theme_bw() +
  labs(title = "# of Cars for Class by Manufacturer",
       x = NULL,
       y = "# of cars",
              ) +
  coord_flip() +
  guides(fill = "none")
```

### 4.4.2. Scatter plot (x: continous, y:continous)

```{r}
mpg %>%
  ggplot(aes(cyl, cty)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + #lm = linear model; se = standard error
  geom_jitter() +
  theme_economist_white() +
  labs(title = "City Mileage versus Cylinders",
       subtitle = "Scatter and jitter plot with regression line",
       caption = "Source: mpg dataset",
       x = "# of Cylinders",
       y = "City mileage",
       )

```

### 4.4.3. x = categorical, y = continous

-   A box plot is best for this situation. A box plot will produce an average of y variable for each level of x.

-   To use a bar plot for this situation, one needs to average y variable per each level of x variable first before drawing a bar plot.

#### 4.4.3.1. Boxplot

```{r tweaking fill and color in boxplot}
ggplot(data=cars, aes(x=cyl, y=mpg, group=cyl)) + # the same as below
  geom_boxplot(aes(fill = cyl)) # different hues of the same color since cyl is continuous

ggplot(data=cars, aes(x=cyl, y=mpg, group=cyl)) + # the same as below
  geom_boxplot(aes(fill = factor(cyl))) # different color since cyl has been factorized and became categorical.


ggplot(data=cars, aes(x=factor(cyl), y=mpg, group = factor(cyl))) + # group argument doesn't add color.
  geom_boxplot()

ggplot(data=cars, aes(x=factor(cyl), y=mpg, fill = factor(cyl))) + # use fill for coloring inside the box.
  geom_boxplot() + 
   labs(x = "# of Cylinder",
       y = "Average Mile Per a Galon of Gasoline",
       title = "Fuel Efficiency by Cylinder Size",
       ) +
  guides(fill = "none")

ggplot(data=cars, aes(x=factor(cyl), y=mpg, color = factor(cyl)))+ # use color to add color to the borderlines.
  geom_boxplot()
```

#### 4.4.3.2. geom_bar(stat = "identity")

::: callout-tip
`geom_bar(stat = "identity")` = `geom_col()`
:::

##### (1) Wrong visualization

```{r geom_bar/geom_col problem}

# when y is not a count but another continuous variable, geom_bar() needs stat = 'identity' as below. In this case, just use geom_col().
ggplot(data=cars, aes(x=factor(cyl), y=mpg, fill = factor(cyl))) +
  geom_bar(stat = 'identity') # the same as below

ggplot(data=cars, aes(x = cyl, y=mpg, fill = cyl)) +
  geom_bar(stat = 'identity') # similar to the above but different hues of the same color.

ggplot(data=cars, aes(x=factor(cyl), y=mpg, fill = factor(cyl))) +
  geom_col() # simpler than geom_bar
```

::: callout-warning
However, notice the mpg in the charts created above. mpg for each level of x variable is too high? Would it be sum or average of y for each level of x? Let's wrangle data first before visualization. You will learn how to wrangle data in another module. Please bear with me for now.
:::

-   **sum**: The result below shows the same y variable as the one for the two charts above.

```{r}
cars %>%
  group_by(cyl) %>% # group_by() is from dplyr
  summarize(mpg_sum = sum(mpg)) %>% # summarize() is from dplyr
  ggplot(aes(x=factor(cyl), y=mpg_sum, fill = factor(cyl))) +
  geom_col()
```

-   **average**: The result below shows the same y variable as the one for boxplot.

```{r}
cars %>%
  group_by(cyl) %>%
  summarize(mpg_mean = mean(mpg), .groups = "drop") %>%
  ggplot(aes(x=factor(cyl), y=mpg_mean, fill = factor(cyl))) +
  geom_col()
```

::: callout-caution
-   After observing the two charts above, we can conclude that geom_bar or geom_col uses a sum of y for each level of x variable, which is not what we want.
-   Thus, to use geom_bar or geom_col for categorical x and continuous y, we need to average y for each level of categorical x first before we use geom_bar() or geom_col.
:::

##### (2) Correct visualization

-   Calculating mean first during wrangling and use stat="identity" in geom_bar()

```{r}
cars %>%
  group_by(cyl) %>%
  summarize(mpg_mean = mean(mpg), .groups = "drop") %>%
  ggplot(aes(x = factor(cyl), y = mpg_mean, fill = factor(cyl))) +
  geom_bar(stat = 'identity') # the same as above chart
```

-   Using stat = "summary" with fun without wrangling

```{r}
cars %>%
  ggplot(aes(factor(cyl), mpg, fill = factor(cyl))) +
  geom_bar(stat = "summary", fun = mean)

cars %>%
  ggplot(aes(factor(cyl), mpg, fill = factor(cyl))) +
  geom_bar(stat = "summary", fun = median)

cars %>%
  ggplot(aes(factor(cyl), mpg, fill = factor(cyl))) +
  geom_bar(stat = "summary", fun = max)
```

##### (3) Exercise with geom_col()

> *Task*: show average of highway mileage by each car class sorted by the performance of the car from best to the worst.

```{r exercise}
head(mpg)

#corect
mpg %>%
  group_by(class) %>%
  summarise(hwy_mean = mean(hwy)) %>%
  ggplot(aes(x = reorder(class, hwy_mean), y = hwy_mean, fill = class)) +
  geom_col() +
  labs(x = "Class",
       y = "AVG. Highway MPG",
       title = "Average Highway Mile Per Gas by Class"
       ) +
  coord_flip() +
  guides(fill = "none")

```

## 4.5. Three variables (x=Categorical, y=continous, moderator = categorical)

> In this case, both boxplots and barplots work well. With barplot, we don't need to calculate average first as R will automatically do so across all levels of the categorical moderator.

### 4.5.1. Boxplot (with facet_grid)

#### (1) Basic with faceting

```{r labeller}
?mtcars

to_string <- as_labeller(c(`0` = "Automatic", `1` = "Manual"))

ggplot(data=cars, aes(x=factor(cyl), y=mpg, fill = factor(cyl))) +
  geom_boxplot() + 
  facet_grid(~ factor(am), labeller = to_string) + #am Transmission (0 = automatic, 1 = manual)
  labs(title = "Impact of Cylinder size on mpg across am",
       x = "Number of Cylinder",
       y = "Average MPG"
       )
```

#### (2) Changing facet label with labeller argument

```{r}
# change facet label
to_string <- as_labeller(c(`0` = "Automatic", `1` = "Manual")) # for numbers, use ``

ggplot(data=cars, aes(x=factor(cyl), y=mpg, fill = factor(cyl))) +
  geom_boxplot(show.legend = FALSE) +
  facet_grid(~ factor(am), labeller = to_string) +
  labs(title = "Impact of Cylinder size on mpg across am",
       x = "Number of Cylinder",
       y = "Average MPG"
       )
```

### 4.5.2. Barplot

> **Note** We know barplot doesn't work when y is continous, but it works okay when we cut it by another categorical variable (moderator) as you will see below.

#### 4.5.2.1. Dodged Barplot (Changing fill labels)

##### (1) Using mutate()

```{r using mutate}
# change moderator's factor level prior to visualization

mtcars %>% as_tibble() %>%
  mutate(am = factor(am), # change the labels of the factor levels in wrangling
         am = case_when(am == "0" ~ "Automatic",
                        am == "1" ~ "Manual"
                        )
         ) %>%
  ggplot(aes(x = factor(cyl), y = mpg, fill = am)) +
  geom_col(position = "dodge") +
  labs(title = "Impact of the number of cylinder on MPG across AM",
       subtitle = "An example of barplot when we don't need to average y variable",
       x = "The number of cylinder",
       y = "Average MPG",
       fill = "Automatic vs. Manual"
       )
```

##### (2) Using scale_fill_discrete(labels = xxx)

```{r}
# change moderator's factor level using scale_fill_discrete()

to_string <- as_labeller(c(`0` = "Automatic", `1` = "Manual"))
ggplot(data = mtcars, aes(x = factor(cyl), y = mpg, fill = factor(am))) +
  geom_bar(stat = 'identity', position = "dodge") + # am Transmission (0 = automatic, 1 = manual)
  scale_fill_discrete(labels = to_string) + # change the labels of the factor levels in visualization stage
  labs(title = "Impact of Cylinder size on mpg across am",
       x = "Number of Cylinder",
       y = "Average MPG",
       fill = "Automatic vs. Manual"
       )

```

### 4.5.2.2. Stacked vs. Filled Barplot

-   Calculate numbers and pass it via stat = "identity" in Barplot

##### (1) Adding % inside Stacked Barplot

```{r adding stat with geom_text}
# stacked bar chart
mpg %>%
  group_by(manufacturer, class) %>%
  tally() %>%
  mutate(percent = n/sum(n), .group = "drop") %>%
  ggplot(aes(x = reorder(manufacturer, n), y = n, fill = class)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(sprintf("%1.1f", percent*100),"%")),
            position = position_stack(vjust = 0.5),
            color = "white") +
  theme_minimal() +
  coord_flip()

# The chart below is identical to the first with a one exception. ordering of manufacturer is dropped here. Compare both charts and see the difference.
mpg %>%
  group_by(manufacturer, class) %>%
  tally() %>%
  mutate(percent = n/sum(n), .group = "drop") %>%
  ggplot(aes(x = reorder(manufacturer, n), y = n, fill = class)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(sprintf("%1.1f", percent*100),"%")),
            position = position_stack(vjust = 0.5),
            color = "white") +
  theme_minimal() +
  coord_flip()
```

##### (2) Adding % inside Filled Barplot

```{r}
#filled bar chart

mpg %>%
  group_by(manufacturer, class) %>%
  tally() %>%
  mutate(percent = n/sum(n)) %>%
  ggplot(aes(x = manufacturer, y = n, fill = class)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(aes(label = paste0(sprintf("%1.1f", percent*100), "%")),
            position = position_fill(vjust = 0.5),
            color = "white"
  ) +
  theme_minimal() +
  labs(title = "Percent of Cars by class across manufacturers",
       x = "Car Brand",
       y = "Percentage",
  ) +
  coord_flip()
```

### 4.5.3. Scatterplot (two continous variable cut by a categorical variable)

```{r}
mpg %>%
  ggplot(aes(x = factor(cyl), y = cty)) +
  geom_point(alpha = .5) +
  geom_smooth(method = "lm", se = FALSE) + # lm = linear model; se = standard error
  geom_jitter() +
  facet_wrap(~ drv) +
  # facet_grid(cols = vars(trans)) +
  theme_economist_white() +
  labs(title = "Impact of Cylinders on City Mileage across Type of Drive Train",
       subtitle = "Scatter and jitter plot with regression line",
       caption = "Source: mpg dataset",
       x = "# of Cylinders",
       y = "City mileage",
       )
```

## 4.6. Five veriavles (x=Continous, y=continous, color, size, facet)

```{r}
mpg %>%
  ggplot(aes(displ, cty)) +
  geom_point(aes(color = drv, size = cyl),
             alpha = 0.5) +
  geom_smooth() +
  scale_size_continuous(range = c(3, 8)) +
  facet_wrap(~ year)

mpg %>%
  ggplot(aes(displ, cty)) +
  geom_point(aes(color = drv, size = cyl),
             alpha = 0.5) +
  geom_smooth(method = lm) +
  facet_wrap(~ year) +
  scale_size_continuous(range = c(3, 8)) +
  labs(x = "Engine Size",
       y = "City MPG",
       title = "Fuel Efficiency") +
  theme_bw()

```

# 5. Correlations with ggpairs()

## 5.1. Simple chart

```{r ggpairs}
# simple
ggpairs(cars %>% 
          select(cyl, mpg, disp, wt, qsec)
        )

cars %>%
  select(cyl, mpg, disp, wt, qsec) |>
  ggpairs()
```

## 5.2. Correlations gropped by categorical variables

```{r}
ggpairs(cars, columns = 2:5,
        aes(color = factor(am))
        )



ggpairs(iris, columns = 1:4,
        aes(color = Species)
        )
```

# 6. Interactive plots

```{r}
# library(plotly)
plot <- ggplot(data = cars, aes(x = hp, y = qsec)) +
  geom_point() +
  geom_smooth()

ggplotly(plot)

```

# 7. Labeling with ggrepel::geom_label()/ geom_label_repel()

##7.1. Additive method

```{r}
library(ggrepel)
plot + 
  geom_label(aes(label = model))

plot + 
  geom_label_repel(aes(label = model))
```

## 7.2 Piping method

```{r}
cars %>%
  ggplot(aes(x = hp, y = qsec)) +
  geom_point() +
  geom_smooth() +
  geom_label_repel(aes(label = model))
```

# 8. Saving plots

```{r}
ggsave(filename = "mygraph.png", plot, dpi = 300)
```

# 9. References

-   R Studio Cheatsheets: <https://posit.co/resources/cheatsheets/>
-   R4DS Book: <https://r4ds.hadley.nz/data-visualize>
-   ggplot2: elegant graphics for data analysis by Hadley Wickham: <https://ggplot2-book.org/index.html>
-   The R Graph Gallery: <https://r-graph-gallery.com/index.html>
