---
title: "Module 3-1-Principle - Data Visualization with ggplot2 in R"
author: "Angelina Jao"
date: '`r format(Sys.time())`'
format: 
  html: 
    toc: true
    toc-depth: 4
    number-sections: false
    code-line-numbers: true
    code-fold: false
    code-link: true
    embed-resources: true
    df-print: paged
editor: visual
execute: 
  freeze: auto
  warning: false  
  error: true
editor_options: 
  chunk_output_type: console
---

# **Overview**

## **Expected Learning Outcomes**

> After taking this workshop, participants should be able to do following:

1.  Explain the concept of the grammar of graphics when visualizing data with the ggplot2 package.

2.  Be familiar with various types of charts.

3.  Visualize data in counts and proportions.

4.  Select appropriate charts based on strategic considerations (e.g., the characteristics of the data and audience).

5.  Create a chart that involves one or two variables with either categorical or continuous data.

6.  Create a chart by adding a categorical moderator (3rd variable) to the chart involving two or three variables.

7.  Create correlation charts.

8.  Read charts and generate insights.

9.  Describe three popular packages that allow one to visualize data.

10. Explain the concept of the grammar of graphics when visualizing data with the ggplot2 package.

# Loading up the libraries

```{r}
#install.packages('ggrepel')
#install.packages('ggthemes')
#install.packages('scales')
#install.packages('plotly')
#install.packages('lattice')
#install.packages('GGally')
#install.packages("dplyr")
#install.packages("tidyverse")
#install.packages('ggtext')
#install.packages("glue")
#install.packages("gapminder")
#install.packages("ggplot2")
#install.packages("ggrepel")
#install.packages("ggthemes")
#install.packages("tibble")
#install.packages("plotly")
#install.packages("GGally")
#install.packages("tidyverse")
#install.packages("ggtext")
#install.packages("skimr")
#install.packages("magmaR")
#install.packages("lattice")
library(ggplot2) #visualization
library(ggrepel) #labels for data
library(ggthemes) #collections of themes
library(scales) # scale
library(plotly) # interactive chart
library(GGally) # correlation
library(dplyr) # data transformation
library(tidyverse) # mega package containing 8 packages
library(ggtext) # for text visualization
library(glue) # combining multiple component
library(gapminder)
library(tibble)
library(magrittr)
library(skimr)
```

# **1. Understand mtcars data**

## **1.1 Using Help**

```         
?mtcars
```

> A data frame with 32 observations on 11 (numeric) variables.

-   \[, 1\] mpg Miles/(US) gallon

-   \[, 2\] cyl Number of cylinders

-   \[, 3\] disp Displacement (cu.in.)

-   \[, 4\] hp Gross horsepower

-   \[, 5\] drat Rear axle ratio

-   \[, 6\] wt Weight (1000 lbs)

-   \[, 7\] qsec 1/4 mile time

-   \[, 8\] vs Engine (0 = V-shaped, 1 = straight)

-   \[, 9\] am Transmission (0 = automatic, 1 = manual)

-   \[,10\] gear Number of forward gears

-   \[,11\] carb Number of carburetors Note\]

## **1.2 Reading data and converting to a tibble (cars)**

```         
head(mtcars)
```

```{r}
class(mtcars)
mtcars

```

```{r}
cars <- 
  mtcars %>% # piping operator from dplyr (shortcut: Ctrl+Shift+M)
  # everything is connected through piping
    rownames_to_column() %>% # do this before changing the data to tibble as the conversion will remove rownames in tibble.
    as_tibble() %>%  
    rename(model = rowname) %>% 
    print (n = 20, width = Inf)
```

```{r}
cars 
```

## 1.3 Simple Descriptive Statistics

```{r}
summary(cars)
glimpse(cars)
skimr::skim(cars)
```

# 2. Basic Plotting Methods in Base R

```{r}
# Using a built in plotting function
hist(cars$disp, breaks = 10)
```

# 3. Lattice package

```{r}
#shift star for special font
library(lattice)
xyplot(mpg ~ wt,cars)
```

# 4. ggplot2

-   We will use ggplot2 - the best tool in the market for data visualization - from now on

## 4.1. Elaborate Example

### Data wrangling

```{r}
cars <- 
  mtcars |>  # piping operator from dplyr (shortcut: Ctrl+Shift+M)
    rownames_to_column() %>% # do this before changing the data to tibble as the conversion will remove rownames in tibble.
    as_tibble() |>   
    rename(model = rowname)
```

### Plotting

```{r}
cars |> 
  count(cyl)

#control shift m for piping operator

# this is cont starting from below for d-2, 2/23

easy_labels <- c("4" = "4 Cylinder Cars",
                 "6" = "6 Cylinder Cars",
                 "8" = "8 Cylinder Cars"
                 )

cars |> 
  mutate(cyl=factor(cyl)) |>
  ggplot(aes(x = mpg, y = disp, color=factor(cyl))) +     
  geom_point(size = 3, 
             color = "black") +   
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ cyl,
             ncol = 1) +
  theme_bw()

```
